---
title: "Project_1"
author: "Byron Jones"
date: "January 19, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Reproducible Research
# Project 1
# Submitted by Byron Jones
```{r}
# Project dataset contains the following variables:
## *steps*:    number of steps taking in a 5-minute interval (missing values are coded as NA)
## *date*:     the date on which the measurement was taken in YYYY-MM-DD format
## *interval*: identifier for the 5-minute interval in which measurement was taken

# The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset

# Required for the assignment:
## 1.Code for reading in the dataset and/or processing the data
## 2.Histogram of the total number of steps taken each day
## 3.Mean and median number of steps taken each day
## 4.Time series plot of the average number of steps taken
## 5.The 5-minute interval that, on average, contains the maximum number of steps
## 6.Code to describe and show a strategy for imputing missing data
## 7.Histogram of the total number of steps taken each day after missing values are imputed
## 8.Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
## 9.All of the R code needed to reproduce the results (numbers, plots, etc.) in the report
```
# Set working directory, read in the dataset and look at first few rows
```{r}
# set working directory
setwd("//Phchbs-s3159/jonesby1$/data/Byron/R/Reproducible Research")

# read in the dataset and look at first few rows
activity<-read.csv("activity.csv",head=TRUE)
print(head(activity))
```
# Identify days of the week and add this new variable to existing activity list
```{r}
week.days<-weekdays(as.Date(unlist(activity$date)))
activity[["day"]]<-week.days
names(activity)<-c("steps", "date","interval", "day")
```
# Calculate mean total number of steps taken each day and mean total number of steps in each 5-minute interval using the aggregate function
## Histogram of total number of steps taken each day
```{r}
## aggregate data for each day
day.mean<-as.numeric(unlist(aggregate(activity$steps, by=list(activity$date), FUN=sum, na.rm=TRUE)[2]))

## load library ggplot2
library(ggplot2)
day.mean.df<-data.frame(day.mean)
ggplot(data=day.mean.df, aes(day.mean.df$day.mean)) +  
  geom_histogram(bins=30) +
   labs(title="Histogram for total number of steps taken each day") +
   labs(x="total number of steps per day", y="Frequency")
```
```
```
# Mean and median of total number per day
```{r}
print(mean(day.mean))
print(median(day.mean))
```
# Time series plot of average number of steps taken
```{r}
## calculate mean for each 5-minute interval using the aggregate function
interval.mean<-as.numeric(unlist(aggregate(activity$steps, by=list(activity$interval), FUN=mean, na.rm=TRUE)[2]))

## plotting values for x-axis
interval.values<-unique(activity$interval)
## create a data frame for use in ggplot2
interval.df<-data.frame(cbind(interval.values,interval.mean))

## time series plot using ggplot2
ggplot(interval.df, aes(x=interval.values, y=interval.mean)) +
  geom_line() +
  ggtitle("time series plot for 5-minute intervals averaged over all days") +
  labs(x="interval label") +
  labs(y="mean over 5-minute interval")
```
```
```
# Maximum mean number of days
```{r}
## maximum mean number of steps
max.mean<-max(interval.mean)
print(max.mean) 
```
# Label of the interval that contains the maximum
```{r}
## interval containing maximum mean
max.interval<-interval.values[interval.mean==max(interval.mean)]
print(max.interval)
```
# Total number of rows with missing data
```{r}
## total number of rows with missing data (NA)
total.NA<-sum(is.na(activity$steps))
print(total.NA)
```
# Imputation of missing data
```{r}
## imputation rule: replace NA with the mean for that day

## calculate the mean for each day of the week
Saturday.mean<-mean(activity$steps[activity$day=="Saturday"],na.rm=TRUE)
Sunday.mean<-mean(activity$steps[activity$day=="Sunday"],na.rm=TRUE)
Monday.mean<-mean(activity$steps[activity$day=="Monday"],na.rm=TRUE)
Tuesday.mean<-mean(activity$steps[activity$day=="Tuesday"],na.rm=TRUE)
Wednesday.mean<-mean(activity$steps[activity$day=="Wednesday"],na.rm=TRUE)
Thursday.mean<-mean(activity$steps[activity$day=="Thursday"],na.rm=TRUE)
Friday.mean<-mean(activity$steps[activity$day=="Friday"],na.rm=TRUE)

## impute missing data
## copy original dataset
activity.imp<-activity
activity.imp$steps[is.na(activity.imp$steps) & activity.imp$day=="Saturday"]<-Saturday.mean
activity.imp$steps[is.na(activity.imp$steps) & activity.imp$day=="Sunday"]<-Sunday.mean
activity.imp$steps[is.na(activity.imp$steps) & activity.imp$day=="Monday"]<-Monday.mean
activity.imp$steps[is.na(activity.imp$steps) & activity.imp$day=="Tuesday"]<-Tuesday.mean
activity.imp$steps[is.na(activity.imp$steps) & activity.imp$day=="Wednesday"]<-Wednesday.mean
activity.imp$steps[is.na(activity.imp$steps) & activity.imp$day=="Thursday"]<-Thursday.mean
activity.imp$steps[is.na(activity.imp$steps) & activity.imp$day=="Friday"]<-Friday.mean
```
# Recalculate using imputed data set and compare with results obtained when missing data were not imputed
```{r}
## calculate total number of steps taken each day using the aggregate function
day.mean.imp<-as.numeric(unlist(aggregate(activity.imp$steps, by=list(activity.imp$date), FUN=sum, na.rm=TRUE)[2]))
```
# Histogram of total number of steps after imputation
```{r}
day.mean.imp.df<-data.frame(day.mean.imp)
ggplot(data=day.mean.imp.df, aes(day.mean.imp.df$day.mean.imp)) +  
  geom_histogram(bins=30) +
   labs(title="Histogram for total number of steps taken each day after imputation") +
   labs(x="total number of steps per day", y="Frequency")
```
```
```
# Identify weekends and add this identifier to the existing activity list
```{r}
weekend<-(activity.imp$day=="Saturday" | activity.imp$day=="Sunday")
## add this to existing activity list
activity.imp[["daytype"]]<-weekend
activity.imp$daytype[activity.imp$daytype]<-"weekend"
activity.imp$daytype[activity.imp$daytype==FALSE]<-"weekday"
```
# Split data into week days and weekend days (separate datasets)
```{r}
activity.weekday<-activity.imp[activity.imp$daytype=="weekday",]
activity.weekend<-activity.imp[activity.imp$daytype=="weekend",]

# x-axis values for plotting each type of day
interval.values.weekday<-unique(activity.weekday$interval)
interval.values.weekend<-unique(activity.weekend$interval)

## calculate mean for each 5-minute interval using the aggregate function
## for weekdays
interval.mean.weekday<-as.numeric(unlist(aggregate(activity.weekday$steps, by=list(activity.weekday$interval), FUN=mean, na.rm=TRUE)[2]))
## for weekends
interval.mean.weekend<-as.numeric(unlist(aggregate(activity.weekend$steps, by=list(activity.weekend$interval), FUN=mean, na.rm=TRUE)[2]))

## make data frames for ggplot2
interval.weekday.df<-data.frame(cbind(interval.values.weekday,interval.mean.weekday))
interval.weekend.df<-data.frame(cbind(interval.values.weekend,interval.mean.weekend))
```
# Panel plot
```{r}
## weekdays pane
p1<-ggplot(interval.weekday.df, aes(x=interval.values.weekday, y=interval.mean.weekday)) +
  geom_line() +
  ggtitle("time series plot for week days") +
  labs(x="interval label") +
  labs(y="mean over 5-minute interval")
## weekends panel
p2<-ggplot(interval.weekend.df, aes(x=interval.values.weekend, y=interval.mean.weekend)) +
  geom_line() +
  ggtitle("time series plot for weekend days") +
  labs(x="interval label") +
  labs(y="mean over 5-minute interval")
```
# Function multiplot
```{r}

##########################
# Multiple plot function
##########################
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)
  
  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)
  
  numPlots = length(plots)
  
  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                     ncol = cols, nrow = ceiling(numPlots/cols))
  }
  
  if (numPlots==1) {
    print(plots[[1]])
    
  } else {
    
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))
    
    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))
      
      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
```
# Panel plot via function multiplot comparing weekdays and weekends
```{r}
multiplot(p1, p2, cols=1)
```